---
title: "Workshop2.Rmd"
author: "Libbie"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# WORKSHOP 2- Using ggplot2 for communication#


# Adding labels
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se.e = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

# 'subtitle' adds additional detail in a smaller font beneath the title and caption adds text at the bottom right of the plot 
# 'caption' adds text at the bottom right of the plot, often used to describe the source of the data.


```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

# Use labs() to replace axis labels and legend titles.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```
# Annotations
# Can use geom_text() to add text directly into plot to add emphasis
# Using filter to pull data values out from the data itself and include in the plot
```{r}
library(dplyr)
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```
# Scales

```{r}
ggplot(mpg, aes(displ, hwy)) + # ggplot2 automatically adds scales for you
  geom_point(aes(colour = class))
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuouos() +
  scale_y_continuouss() +
  scale_colour_discrete()
```

```{r}
# Create the plot with fixed scale functions and axis limits
# scale_continuous allows you to manual adjust scale, ggplot2 normally would do this automatically
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous(limits = c(1, 7)) +  # Set x-axis range 
  scale_y_continuous(limits = c(10, 45)) +  # Set y-axis range 
  scale_colour_discrete()
```
# 3.4 Axis Ticks
# breaks controls the position of the ticks

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```
# use NULL to suppress labels, sometimes needed for patocular formats
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```
# 3.5 Legends and colour schemes
# Can use theme() to change the position of your legend, perhaps to make the plot itself as wide as possible (e.g. put the legend underneath) or to hide it all together
# can also use legend.position = "none" to suppress the display of the legend altogether
# Each line was run seperatly and then all together to show differences

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default
```
# 3.6 Replacing a scale
# Continuous position scales and colour scales are the most likely to be switched out

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```
# Changing colour scale

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```
# Adding redundant shape mapping

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```
# ColorBrewer scales are documented here http://colorbrewer2.org/
# Setting preferred colors with scale_colour_manual()

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```
# Viridis colour scheme is one of the most popular

install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)

```{r}
df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```
# Wes Anderson colour pallete is popular as well

# 3.7 Themes
# ggplot2 has 8 themes included, more are here https://github.com/jrnold/ggthemes
  # ggthemes add-on package
  
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()
```
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
```
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```


```{r}
 theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial")) 
```

# 3.8 Saving and exporting your plots
#  ggsave() will save the most recent plot to your working directory. 

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
```

# To adjust dimensions of plot, use width and the length arguments to funciton this

