---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Ben Cresswell"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Install and load tidyverse packages
```{r}
# Delete this line once installed # Deleted line of installing package
library("tidyverse")
```


# Load the data (mpg is built into ggplot2)
```{r}
data(mpg)
```


# Quick data checks
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# Create first plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

# The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?


# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot()
# Or with just the data specified?
ggplot(data = mpg)
```


We need to map some aesthetics!!

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) #added random colour to classes on graph
```


# Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

# Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note the warning!!!


# Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #alpha means shading to plots
```
# Another warning!!
# Question: When would using alpha (or size be appropriate??)
## when dealing with overlapping, desnity size or adding emphasis

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


# Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") #makes entire class blue
```


# Troubleshooting 






# Faceting and panel plots

## facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + #make sure + is on the top line
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2, ncol = 4) # faceting will put each class into its own plots
```


## facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) #removes classes and puts everything into a grid
#Creates multiple subplots to compare how the relationship between displacement (displ) and highway MPG (hwy) varies across drv (drive type) and cyl (cylinders).
#Removes overlapping data by distributing points across small multiples.
#Allows easy pattern detection
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


## facet_grid by just row (or column)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively
# removes hwy, puts everything into rows
```

# Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 
# nrow – Sets the number of rows 
# ncol – Sets the number of columns 
# Other options:
# scales – Controls whether axes are fixed or free.
# dir – Sets facet wrapping direction ("h" for horizontal, "v" for vertical).
# strip.position – Changes label placement ("top", "bottom", "left", "right").
# as.table – Alters facet ordering (TRUE = left-to-right, FALSE = bottom-to-top).


# Lines
We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displx, y = hwy))
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

# To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
### Question: whey don't we use geom_line() here? What would that look like? ###
# geom_point() is used when plotting discrete observations.
# geom_line() is best for continuous time-series or sequential data.
# Using geom_line() incorrectly with categorical data creates misleading visuals. (zig-zag lines)


# So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

## Using comments (#)
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.

# Changing linetype
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result
# different line types for objects and classes
```


# Grouping
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
 
# Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```

# Multiple geoms 
# Plotting mulitple geoms on a single plot
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```


# Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth() #layer aesthetic, points are styled by themselves
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!


```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE) # specify different data for each layer
# "subcompact" filter is used to select a subset of the dara and plot only that subset
```
# Question: what does the "se = FALSE" do? 
# Removes the confidence interval (shaded area) around the smoothed line in geom_smooth(), displaying only the trend line.


# Exercise:
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

# Line chart → geom_line()
# Boxplot → geom_boxplot()
# Histogram → geom_histogram()
# Area chart → geom_area()



3.  Will these two graphs look different? Why/why not?
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```


# Transformations and stats

## Load data
```{r}
data("diamonds")
glimpse(diamonds)
```
# The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation. 
# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis?

# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

Question: does anyone get the warning that's in the workbook?
# Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
# If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))

# Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
# Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)
# Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


# Filling by a variable
# Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


# To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


# Position adjustments
# position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


# position = "dodge" 
#Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering
# position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
  geom_point(mapping = aes(x = displ, y = hwy))
```

# Plot Deconstrcution

# Load necessary libraries
library(ggplot2)
library(readr)

# Step 1: Load the extracted data
# If you exported as CSV from WebPlotDigitizer, load it:
 # Replace with actual file path
```{r}

getwd()
data <- read_csv("../data/Default Dataset.csv")
```
```{r}
# Assign meaningful column names
colnames(data) <- c("Year", "MLD")

# Check the structure again
glimpse(data)
```
```{r}
# Create an improved line plot
ggplot(data, aes(x = Year, y = MLD)) + 
  geom_line(color = "black", linewidth = 1) +  # Main trend line
  geom_point(color = "black", linewdith = 2) +  # Data points
  labs(title = "Improved Visualization of MLD Over Time",
       x = "Year",
       y = "Mixed Layer Depth (MLD in meters)") +
  theme_minimal() +  # Clean background
  theme(
    text = element_text(size = 14),  # Increase font size
    axis.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "gray90")  # Light grid lines
  )
```

##TAKE TWO##

```{r}
getwd()
data <- read_csv("../data/wpd_datasets.csv")
```
```{r}
# Assign meaningful column names
colnames(data) <- c("Year", "MLD", "Lower_CI", "Upper_CI", "Lower_CI_Extra", "Upper_CI_Extra")

data <- data %>%
  mutate(across(where(is.character), as.numeric))

# Check the structure again
glimpse(data)
```

```{r}
ggplot(data, aes(x = Year, y = MLD)) + 
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI), fill = "gray", alpha = 0.3) +  # Proper confidence interval
 #geom_line(color = "black", size = 1) +  # Main trend line
  geom_point(color = "black", size = 2) +  # Data points
  labs(title = "Improved Visualization of MLD Over Time with Confidence Interval",
       x = "Year",
       y = "Mixed Layer Depth (MLD in meters)") +
  theme_minimal() +  # Clean background
  theme(
    text = element_text(size = 14),  # Increase font size
    axis.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "gray90")  # Light grid lines
  )

```











