---
title: "Workshop4.Rmd"
author: "Libbie"
date: "2025-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load the copepod data into R studio

```{r}
library(readr)
```


```{r}
#load the copepod data into R studio
library(readr)
dat <- read_csv("../data/data-for-course/copepods_raw.csv")
dat
```

```{r}
library(ggplot2) #  plotting the coordinates for the samples (segments of the CPR silks)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
# visualizing richness in a non-spatial way with latitude on the x-axis and richness on the y-axis. 

```
# 5.8
# st_as_sf converts different data types to simple features. 
dat is our original data. 
# coords gives the names of the columns that relate to the spatial coordinates (in order of X coordinate followed by Y coordinate).
# crs stands for coordinate reference system which we will discuss next.

# Turning data into a ‘simple features collection’.(sf)


```{r}
install.packages("sf")
library("sf")
```
```{r}
str(dat)
head(dat)

sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), #turn our data into a ‘simple features collection’.
                 crs = 4326)
```

# 5.9 Coordinate reference systems
#coordinate reference systems are required for 2D mapping to compensate for the lumpy, spherical (3D) nature of the earth.


```{r}
crs4326 <- st_crs(4326)
crs4326 # look at the whole CRS
crs4326$Name ## pull out just the name of the crs
crs4326$wkt # crs in well-known text format
```

# 2D mapping^

5.10 Feature collection (points)


```{r}
sdat #sf is that because the data is basically a dataframe with a geometry, we can use all the operations that work on dataframes on sf simple features collections.

```


# 5.11 Cartography

```{r}
plot(sdat["richness_raw"])
```

```{r}
plot(sdat)
```


```{r}
plot(sdat["sample_time_utc"])
```

# 5.12 Thematic maps for communication

```{r}
install.packages("tmap")
```

```{r}
library(tmap)
```

```{r}
#using tmap

tm_shape(sdat) + 
  tm_dots(col = "richness_raw")
```

```{r}
tm_shape(sdat) + 
  tm_symbols(col = "richness_raw")
tmap_save()
```
```{r}
# Assuming 'sdat' is your spatial data object and 'richness_raw' is a column in 'sdat'
library(tmap)
tm1 <- tm_shape(sdat) + 
  tm_dots(col = "richness_raw")

```

```{r}
tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600)
```

# 5.13 Mapping spatial polygons as layers

# We can read shapefiles directly into R with the st_read command (which is like read_csv, but for spatial files)

```{r}
aus <- st_read("../data/data-for-course/spatial-data/Aussie/Aussie.shp")
```

```{r}
shelf <- st_read("../data/data-for-course/spatial-data/aus_shelf/aus_shelf.shp")
```

```{r}
aus
```
# 5.13.2 Mapping your polygons

```{r}
tm_shape(shelf) + 
  tm_polygons()
```

```{r}
tm_shape(shelf, bbox = sdat) + # bbox = sdat to expand the extent of the map so it depicts all of our copepod data points. 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots()
```
# 5.14 Exploring t_map

```{r}
install.packages("tmap", dependencies = TRUE)
```








